Type-Safe Modular Parsing

### 2016-12-29 paper/sections/Inheritance.tex
First part of Object Algebras.

\subsection{Dimensions of Extensibility}
We continue with the expression language of literals and additions. At this point,
one would like to have two kinds of extensions:

@@

\lstinputlisting[linerange=95-97]{../Scala/Parser/src/PaperCode/Overview/Overview.scala}% APPLY:linerange=OVERVIEW_SIMPLE_LAM

@@

Although delegation enables OO extensibility, the use of an OO class
hierarchy makes the addition of new operations over the AST
problematic. Modification on existing code breaks the modularity. This
refers to the famous \textit{Expression Problem} []. For challenge 2, suppose now we want to
support evaluation on expressions, in Scala one attempt would be extending \lstinline{Expr} with
evaluation:

\lstinputlisting[linerange=178-178]{../Scala/Parser/src/PaperCode/Overview/Overview.scala}% APPLY:linerange=OVERVIEW_ATTEMPT_EXPRWITHEVAL
But then the programmer has to define new classes for literals and additions that extend old ones and \lstinline{EvalExpr}. Specifically,
the critical point is that existing parsing code only generates \lstinline{Expr} objects with pretty printing, so modification on parsing is again required.

In contrast, Object Algebras~\cite{Oliveira2012} enable us to solve this problem, as it separates data variants and
operations, to
offer high flexibility in the choice of operations to be performed
over the AST. It also makes parsing with multiple sorts of syntax easier.

\paragraph{Parsing with Object Algebras} Using Object Algebras, the abstract syntax of the expression language is defined as below.

\lstinputlisting[linerange=8-11]{../Scala/Parser/src/PaperCode/Overview/OA.scala}% APPLY:linerange=OVERVIEW_OA_ALG

Then we are able to define operations over the syntax in a modular way. For instance, pretty printing operation can be realized as:

\lstinputlisting[linerange=15-18]{../Scala/Parser/src/PaperCode/Overview/OA.scala}% APPLY:linerange=OVERVIEW_OA_PRINT

@@

\lstinputlisting[linerange=22-27]{../Scala/Parser/src/PaperCode/Overview/OA.scala}% APPLY:linerange=OVERVIEW_OA_PARSER

@@

\lstinputlisting[linerange=31-44]{../Scala/Parser/src/PaperCode/Overview/OA.scala}% APPLY:linerange=OVERVIEW_OA_EXT
Note that we use the new combinator \lstinline{<|>} defined in our library, to compose two parsing functions into one. The function \lstinline{pExtExpr} has the compound type \lstinline{ExprAlg[E] with VarAlg[E]} in its argument, where two language ASTs are combined to represent the new language.

@@

\lstinputlisting[linerange=53-58]{../Scala/Parser/src/PaperCode/Overview/OA.scala}% APPLY:linerange=OVERVIEW_OA_USE

@@

\lstinputlisting[linerange=72-78]{../Scala/Parser/src/PaperCode/Overview/OA.scala}% APPLY:linerange=OVERVIEW_OA_EXT_OP

@@

\lstinputlisting[linerange=63-67]{../Scala/Parser/src/PaperCode/Overview/OA.scala}% APPLY:linerange=OVERVIEW_OA_MULTI_SYNTAX
In the next sections, we will discuss them with more details.

>>>>

\subsection{Dimensions of Extensibility}\label{subsec:dimensions}
We continue with the expression language of literals and additions. At this point,
one would like to have two kinds of extensions:

@@

\lstinputlisting[linerange=98-100]{../Scala/Parser/src/PaperCode/Overview/Overview.scala}% APPLY:linerange=OVERVIEW_SIMPLE_LAM

@@

@@

\lstinputlisting[linerange=29-34]{../Scala/Parser/src/PaperCode/Overview/OA.scala}% APPLY:linerange=OVERVIEW_OA_PARSER

@@

\lstinputlisting[linerange=38-51]{../Scala/Parser/src/PaperCode/Overview/OA.scala}% APPLY:linerange=OVERVIEW_OA_EXT
Note that we use the new combinator \lstinline{<|>} defined in our library, to compose two parsing functions into one. The function \lstinline{pExtExpr} has the compound type \lstinline{ExprAlg[E] with VarAlg[E]} in its argument, where two language ASTs are combined to represent the new language.

@@

\lstinputlisting[linerange=60-65]{../Scala/Parser/src/PaperCode/Overview/OA.scala}% APPLY:linerange=OVERVIEW_OA_USE

@@

\lstinputlisting[linerange=79-85]{../Scala/Parser/src/PaperCode/Overview/OA.scala}% APPLY:linerange=OVERVIEW_OA_EXT_OP

@@

\lstinputlisting[linerange=70-74]{../Scala/Parser/src/PaperCode/Overview/OA.scala}% APPLY:linerange=OVERVIEW_OA_MULTI_SYNTAX
In the next sections, we will discuss them with more details.

### 2016-12-29 paper/sections/ObjectAlgebras.tex
Added S3, inheritance-based approach.
First part of Object Algebras.

\section{Object Algebras and Parsing}\label{sec:algebrasandparsing}

Object Algebras~\cite{Oliveira2012} is a solution to the famous \textit{Expression Problem}, provide extensibility on
both data variants and operations. It is suitable for structures like abstract syntax trees. We integrate Object Algebras directly to enhance the extensibility of \name.

@@

In Object Algebras, ASTs as recursive data structures are defined using traits, where each constructor corresponds
to an abstract method inside. The example from Section~\ref{subsec:parsingwithopen} is again used here for illustration.
At first the language only supports variables and applications:

\lstinputlisting[linerange=4-8]{../Scala/Parser/src/PaperCode/ObjectAlg/OA1.scala}% APPLY:linerange=OBJECT_ALG_EXPR

The language has two constructors: literals and additions. The trait \lstinline{ExprAlg} is called an \textit{Object Algebra interface},
whereas as a factory, it cannot produces objects for the expressions directly like traditional approaches, but abstract the results in its
type parameter \lstinline{E} instead. To realize an operation on expressions, we simply instantiate the type parameter by a concrete type and
provides implementations for all cases. Below is an example of collecting all free variables in an expression:

\lstinputlisting[linerange=12-15]{../Scala/Parser/src/PaperCode/ObjectAlg/OA1.scala}% APPLY:linerange=OBJECT_ALG_FREE_VARS

Here \lstinline{FreeVars} is called an \textit{Object Algebra}. It traverses an expression bottom-up, and returns a list of strings as the result.
Furthermore, one can define a new trait to implement pretty-printing on expressions:

\lstinputlisting[linerange=19-22]{../Scala/Parser/src/PaperCode/ObjectAlg/OA1.scala}% APPLY:linerange=OBJECT_ALG_EXPR_PRINT

On the other hand, the data variants can be extended by inheriting \lstinline{ExprAlg} and adding new cases only. Suppose we want to
have literals in expressions, a new Object Algebra interface \lstinline{LitAlg} is defined as follows:

\lstinputlisting[linerange=26-28]{../Scala/Parser/src/PaperCode/ObjectAlg/OA1.scala}% APPLY:linerange=OBJECT_ALG_LIT

@@

\lstinputlisting[linerange=32-34]{../Scala/Parser/src/PaperCode/ObjectAlg/OA1.scala}% APPLY:linerange=OBJECT_ALG_LIT_PRINT

@@

\huang{todo: better name for build}
\lstinputlisting[linerange=48-48]{../Scala/Parser/src/PaperCode/ObjectAlg/OA1.scala}% APPLY:linerange=OBJECT_ALG_BUILD

@@

\lstinputlisting[linerange=38-42]{../Scala/Parser/src/PaperCode/ObjectAlg/OA1.scala}% APPLY:linerange=OBJECT_ALG_REFACTOR

>>>>

\section{Extensible ASTs with Object Algebras}\label{sec:algebrasandparsing}

The inheritance-based approach benefits us from building extensible parsers, based on an OO class hierarchy.
Nevertheless, the addition of new operations over the AST becomes problematic. Often it requires modification on
the existing code, which breaks modularity. In this section, we will have a closer look at how it motivates us to
use Object Algebras for full extensibility.

\subsection{Problem with Traditional OO ASTs}\label{subsec:problemwithoutoa}

Now it comes to extension 2 in Section~\ref{subsec:dimensions}. In the previous code, we have presented pretty-printing as
an example of operations on the AST. Suppose now we also want to collect free variables on the expression language, which models
literals, additions and variables. Based on the traditional OO class hierarchy, one attempt would be extending \lstinline{Expr} with the new
operation directly:

\lstinputlisting[linerange=89-89]{../Scala/Parser/src/PaperCode/Overview/Overview.scala}% APPLY:linerange=ATTEMPT_EXPRWITHFREEVARS
It refers to one Scala solution to the Expression Problem []. But then programmers have to define new classes for literals, additions and variables
by extending the old ones together with \lstinline{FreeVarsExpr}. The critical issue is that, the existing parsing code only generates objects of the old
types, again it is a modularity issue, since it is a bad idea to modify existing code again and again with extensions. Other solutions including modelling constructors as functions introduce unnecessary complexity.

In contrast, Object Algebras~\cite{Oliveira2012} enables us to solve this problem. As a nice solution to the Expression Problem, it separates data variants
and operations, and offers high flexibility in the choice of operations to be performed over the AST. Before adopting it for parsing, the next section introduces the background of Object Algebras.

@@

In Object Algebras, ASTs as recursive data structures are defined using traits, where each constructor corresponds
to an abstract method inside. The example from Section~\ref{subsec:packratparsing} is again used here for illustration.
At first the language only supports literals and additions:

\lstinputlisting[linerange=8-11]{../Scala/Parser/src/PaperCode/Overview/OA.scala}% APPLY:linerange=OVERVIEW_OA_ALG

The language has two constructors: literals and additions. The trait \lstinline{ExprAlg} is called an \textit{Object Algebra interface},
whereas as a factory, it cannot produces objects for the expressions directly like traditional approaches, but abstract the results in its
type parameter \lstinline{E} instead. To realize an operation on expressions, we simply instantiate the type parameter by a concrete type and
provides implementations for all cases. Below is an example of pretty-printing an expression:

\lstinputlisting[linerange=15-18]{../Scala/Parser/src/PaperCode/Overview/OA.scala}% APPLY:linerange=OVERVIEW_OA_PRINT

Here \lstinline{Print} is called an \textit{Object Algebra}. It traverses an expression bottom-up, and returns a string as the result.
Furthermore, one can define a new trait to implement evaluation on arithmetic expressions:

\lstinputlisting[linerange=18-21]{../Scala/Parser/src/PaperCode/ObjectAlg/OA1.scala}% APPLY:linerange=OBJECT_ALG_EXPR_PRINT

On the other hand, the data variants can be extended by inheriting \lstinline{ExprAlg} and adding new cases only. Now we extend the language
with variables. A new Object Algebra interface \lstinline{VarAlg} is defined as follows:\haoyuan{Updates till here.}

\lstinputlisting[linerange=25-27]{../Scala/Parser/src/PaperCode/ObjectAlg/OA1.scala}% APPLY:linerange=OBJECT_ALG_LIT

@@

\lstinputlisting[linerange=31-33]{../Scala/Parser/src/PaperCode/ObjectAlg/OA1.scala}% APPLY:linerange=OBJECT_ALG_LIT_PRINT

@@

\huang{todo: better name for build}
\lstinputlisting[linerange=47-47]{../Scala/Parser/src/PaperCode/ObjectAlg/OA1.scala}% APPLY:linerange=OBJECT_ALG_BUILD

@@

\lstinputlisting[linerange=37-41]{../Scala/Parser/src/PaperCode/ObjectAlg/OA1.scala}% APPLY:linerange=OBJECT_ALG_REFACTOR

### 2016-12-29 paper/sections/PackratParsing.tex
First part of Object Algebras.

\subsection{Packrat Parsing}
Fortunately some more advanced parsing techniques such as Packrat parsing~\cite{Ford2002}, address the limitations of simple parser combinators such as Parsec. Packrat parsers use memoization to record the result of applying each parser at each position of the input, so that repeated computation is eliminated.  Moreover, theoretically the algorithm behind Packrat parsers supports both direct and indirect left-recursion~\cite{warth2008}. All of these properties are very suitable for modularity, thus we decided to use Packrat parsers as the underlying parsing technique in our modular parsing library.
\begin{comment}
It is worth mentioning that the choice of parser combinators will not
affect the other parts of our library. One can choose other parser
combinators like Parsec, in cases that the performance and supporting
of left-recursion are not major concerns. A different library can even build a new
\name with fancy features or higher efficiency.
\end{comment}
In the current version of Scala (2.11), Packrat Parsing is included
in the Scala standard parser combinator library. The library provides a number of
parser combinators, together with some basic parsers. Below we present an example to illustrate Scala parsers.
%
%\bruno{So, we make a big fuss identifying the problems with other
%parser combinators in the previous subsection and now we present
%an example that does not suffer from any of the previous problems!
%What makes sense here is an example that uses: a \emph{left-recursive grammar}; the \emph{longest match combinator}; and where we can argue that
%there's no need for the user to do manual backtracking. Just use
%the example with the left-recursive grammar for parsing integers and additions and argue that none of the limitations
%discussed previously applies!
%}

@@

It is straightforward to model ASTs by inheritance and write corresponding parsers for all cases.
Note that a parser has type \lstinline{PackratParser[E]} for some
\lstinline{E}, which indicates the type of results it produces. Figure~\ref{fig:packrat-arith} shows the abstract syntax and
the parser code that generates expressions as ASTs.

>>>>

\subsection{Packrat Parsing}\label{subsec:packratparsing}
Fortunately some more advanced parsing techniques such as Packrat parsing~\cite{Ford2002}, address the limitations of simple parser combinators such as Parsec. Packrat parsers use memoization to record the result of applying each parser at each position of the input, so that repeated computation is eliminated.  Moreover, theoretically the algorithm behind Packrat parsers supports both direct and indirect left-recursion~\cite{warth2008}. All of these properties are very suitable for modularity, thus we decided to use Packrat parsers as the underlying parsing technique in our modular parsing library.
\begin{comment}
It is worth mentioning that the choice of parser combinators will not
affect the other parts of our library. One can choose other parser
combinators like Parsec, in cases that the performance and supporting
of left-recursion are not major concerns. A different library can even build a new
\name with fancy features or higher efficiency.
\end{comment}
In the current version of Scala (2.11), Packrat Parsing is included
in the Scala standard parser combinator library. The library provides a number of
parser combinators, together with some basic parsers. Below we present an example to illustrate Scala parsers.
%
%\bruno{So, we make a big fuss identifying the problems with other
%parser combinators in the previous subsection and now we present
%an example that does not suffer from any of the previous problems!
%What makes sense here is an example that uses: a \emph{left-recursive grammar}; the \emph{longest match combinator}; and where we can argue that
%there's no need for the user to do manual backtracking. Just use
%the example with the left-recursive grammar for parsing integers and additions and argue that none of the limitations
%discussed previously applies!
%}

@@

It is straightforward to model ASTs by inheritance and write corresponding parsers for all cases.
Note that a parser has type \lstinline{PackratParser[E]} for some
\lstinline{E}, which indicates the type of results it produces. Figure~\ref{fig:packrat-arith} shows the abstract syntax and
the parser code that generates expressions as ASTs. Meanwhile, pretty-printing shows up as an operation on the AST.
